Assessing landscape N removal in coastal New England catchments using the N-Sink

### 2023-04-18 manuscript/manuscript.Rmd
Incorporate edits from AG and CA

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index. To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England. Users can use the R package to develop N delivery maps for a HUC-12 of interest, or can visit the University of Connecticut website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = FALSE,
  tidy = FALSE,
  message = FALSE
)
#load(system.file("testdata.rda", package = "nsink"))
```

@@

```{r get_huc, include = FALSE, eval = TRUE}
# The Niantic
niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
# The Farm
farm_huc_id <- nsink_get_huc_id("Farm River")$huc_12
```

```{r get_data_norun, eval = TRUE, include = FALSE, cache=TRUE}
# Change eval = TRUE if you want to download the data
# Get data for selected HUC
niantic_download <- nsink_get_data(niantic_huc_id, 
                                   data_dir = "nsink_niantic_data")
farm_download <- nsink_get_data(farm_huc_id, data_dir = "nsink_farm_data")
```

@@

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

Functions|Description
:--------|:----------
nsink_get_huc_id|Takes Hydrologic Unit Code (HUC) Name and returns HUC12 IDs
nsink_get_data|Downloads local copies of datasets for specified HUC
nsink_prep_data|Standardizes projections, clips all datasets to HUC boundary
nsink_calc_removal|Calculates percentage of nitrogen (N) removal within the various landscape N sinks
nsink_generate_flowpath|Generates a flowpath from XY starting point for use in N removal analysis
nsink_summarize_flowpath|Estimates N removal along flowpath and lists removal by type of N sink encountered
nsink_generate_static_maps|Generates static maps for a given HUC
nsink_plot|Creates simple plot with pre-selected palettes from a list of static maps created by nsink_generate_static_maps()
nsink_build|Wrapper function that runs all required functions to build the full dataset and static maps for a specified HUC12

## Using the `nsink` package
The `nsink` package provides several functions to set up and run an N-Sink analysis for a specified 12-digit HUC. By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.  Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC.

The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.

@@

`nsink_get_data()` - download and save in a data folder hydrography, soils, and land cover for the specified 12-digit HUC. These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC. The `nsink` package utilizes openly available data from several U.S. Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.  The HUC ID is required and users may specify a path for storing the data as well as indicate whether or not to download the data again if they already exist in the data directory.

@@

```{r prep_data, eval = TRUE, include = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072

# Prep data for selected HUC
niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea, data_dir = "nsink_niantic_data")
farm_data <- nsink_prep_data(farm_huc_id, projection = aea,
                             data_dir = "nsink_farm_data")
```

@@

```{r calc_removal, eval = TRUE, include = FALSE}
# Calculate removal from prepped data
niantic_removal <- nsink_calc_removal(niantic_data)
```

## Generate and summarize flowpaths
A useful part of the N-Sink approach is not just the calculation of potential N removal for individual components of the landscape, it is the ability to summarize cumulative removal along the length of a specified flowpath. Starting from any specified location the flowpath on land is generated from a flow direction grid.  Once that flowpath intersects the surface water network, flow is determined by flow along the NHD Plus surface water network. With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function. It takes the flowpath and removal as input. A data frame is returned with each segment identified by type of sink, if present (i.e., "Hydric", "Stream", "Lake/Pond", or "No Removal"), the percent removal associated with that segment, and cumulative removal.  Total cumulative removal is 100 - minimum of the `n_out` column. That is, `n_out` keeps track of the percent of N leaving each subsequent downstream segment (Table 2).

```{r summarize_it, message = FALSE, include = FALSE}
library(sf)
## Linking to GEOS 3.8.1, GDAL 3.1.4, PROJ 6.3.1
# Starting point

@@

```
```{r show_summary, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_a, caption = "Table 1: Summary of N removal along specified flowpath A (See Figure 4)")
```
## Static maps

@@

```{r static_maps, eval=TRUE, include = FALSE}
niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
```

@@

```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 100%) is the estimeated percentage of N removed within three types of landscape N sinks - wetlands, streams, ponds."}

nsink_plot(niantic_static_maps, map = "removal")
```
```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) is the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}

nsink_plot(niantic_static_maps, map = "transport")
```
```{r plot3, fig.cap = "Figure 3: Nitrogen Delivery Index (range 0 to 100%)  is a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters. N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1. These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI."}

@@

In estimating N removal occurring in the terrestrial portion of the flowpath, we originally focused on riparian wetlands, using SSURGO mapped hydric soils to identify this landscape sink. With the latest version of SSURGO we include all hydric soils in the catchment, except those identified as "subaqueous". Each soil mapping unit (SMU) that has a positive hydric status also has an estimate of %hydric for that SMU. We then estimate N removal within that SMU as 80% of the %hydric. Each raster has one SMU identified with it and raster cell size is 30m X 30m. A 30-m wide wetland is estimated by Mayer et al. [-@mayer2007meta] as having 80% N removal. In the current version N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.

## Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD data (e.g. time of travel) associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system. However, they can represent more than 5% of the drainage area in xx% of the catchment analyzed for southern New England. We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. A subset of these off-network stream segments are identified by the NHD dataset as ditches and canals. These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment due to the level of uncertainty regarding residence time in these groundwater-fed lakes and ponds [@hampton2019residence].

## Caveats
N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated. In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypass the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated. The variation in loading associated with different agricultural practices, density of unsewered development and inputs from wastewater treatment plants are not encompassed in the default loading index. Thus the numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations. It is not intended to replace high-resolution models that rely on long term weather records and incorporate detailed spatial data on N loading, within-field transport or variable source hydrology (e.g., Easton [-@easton2008re]). Instead, N-Sink combines widely available datasets for waterway networks, soils and land cover with nitrogen dynamics developed for these datasets to highlight major sources and sinks of nitrogen within a watershed context.

@@

Users can use the following to interactively identify a starting point for a flowpath of interest. 
```{r case1, echo = TRUE, eval = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072
# Select location on map for starting point
pt <- unlist(locator(n = 1))
# Convert to sf POINT
start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
# Generate the flowpath
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
# Summarize N removal along the flowpath
niantic_removal <- nsink_calc_removal(niantic_data)
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) # table of removal within flowpath segments
```

Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters. This example illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Figure 4). The difference is due to the different N sinks encountered along the way to the estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3). These two undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet. If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing N transformation.

```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations with differing N removal estimates. Hydrofeatures are pictured on the  Nitrongen Transport Index map."}

@@

niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
                     "#F9C159", "#F99844", "#F56A2E","#EF2820"))
plot(niantic_data$streams, add = TRUE, col = "dodgerblue")
plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
ab_plot <- st_sf(st_sfc(ab), crs = aea)
plot(ab_plot, add = TRUE, pch = 16)
text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
```

```{r compare_flowpaths, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
```

@@

The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration. For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters. Variances granted on a case-by-case basis can quickly undermine local water quality goals. Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize acquisition or wetland restoration projects.

@@

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest, highlighting two developed areas."}
box <- c(1949902, 1953930, 2288202, 2292262)
e <- extent(box)
legend_nlcd <- pal_nlcd()
case2 <- crop(niantic_data$nlcd, e)
vals <- unique(case2@data@values)
leg <- legend_nlcd[legend_nlcd$ID %in% vals,]
case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))

my_key <- list(rectangles=list(col = leg$Color),
              text=list(lab=leg$Class),
              space='left',
              columns=1,
              size=2,
              cex=.6)
colors <- legend_nlcd$Color
plot(case2, col = legend_nlcd$Color)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
```
```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery as estimated based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
plot(case2, col = legend_nlcd$Color, alpha = 0.6, legend = FALSE)
plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)

>>>>

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to understand the effects of their decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index. To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England. The R package can be used to develop N delivery maps for a HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = FALSE,
  tidy = FALSE,
  message = FALSE
)
#load(system.file("testdata.rda", package = "nsink"))
#load("testdata.rda")
```

@@

```{r get-huc, include = FALSE, eval = TRUE}
# The Niantic River
niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
```

```{r get-data, eval=TRUE, cache=FALSE, include=FALSE}
# eval = TRUE if you want to download the data
# Get data for selected HUC
niantic_download <- nsink_get_data(niantic_huc_id, 
                                  data_dir = "nsink_niantic_data")
```

@@

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

## Using the `nsink` package
The `nsink` package provides several functions (Table 1) to set up and run an N-Sink analysis for a specified 12-digit HUC. By using the package all required data are downloaded, prepared for the analysis, HUC-wide N removal calculated, and flowpaths summarized.  Additionally, a convenience function, `nsink_build()`, is included that will run all of the required functions for a specified HUC. The workflow follows a simple series of steps to create a set of static maps: N Removal Efficiency, N Transport Efficiency, and N Delivery Index.

```{r table1-build, eval = TRUE, include = FALSE}
table1 <- data.frame(`Function` = c("nsink_get_huc_id", "nsink_get_data", 
                                  "nsink_prep_data", "nsink_calc_removal",
                                  "nsink_generate_flowpath", "nsink_summarize_flowpath",
                                  "nsink_generate_static_maps", "nsink_plot",
                                  "nsink_build"),
                     `Description` = c("Takes Hydrologic Unit Code (HUC) Name and returns HUC12 IDs",
                                       "Downloads local copies of datasets for specified HUC",
                                       "Standardizes projections, clips all datasets to HUC boundary",
                                       "Calculates percentage of nitrogen (N) removal within the various landscape N sinks",
                                       "Generates a flowpath from XY starting point for use in N removal analysis",
                                       "Estimates N removal along flowpath and lists removal by type of N sink encountered",
                                       "Generates static maps for a given HUC",
                                       "Creates simple plot with pre-selected palettes from a list of static maps created by nsink_generate_static_maps()",
                                       "Wrapper function that runs all required functions to build the full dataset and static maps for a specified HUC12"))

```
```{r table1, include = TRUE, echo = FALSE}
knitr::kable(table1, caption = "Table 1: Functions in the R package `nsink`")
```

@@

`nsink_get_data()` - download and save in a data folder hydrography, soils, and land cover for the specified 12-digit HUC. These data cover the HUC, but have not been pared down to those data exclusive to the specified HUC. The `nsink` package utilizes widely available data from several U.S. Federal Government sources and as of `r lubridate::today()` no authentication is required to access these sources.  The HUC ID is required and users may specify a path for storing the data as well as indicate whether or not to download the data again if they already exist in the data directory.

@@

```{r prep-data, eval = TRUE, include = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072

# Prep data for selected HUC
niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea,
                                data_dir = "nsink_niantic_data")
```

@@

```{r calc-removal, eval = TRUE, include = FALSE}
# Calculate removal from prepped data
niantic_removal <- nsink_calc_removal(niantic_data)
```

## Generate and summarize flowpaths
A useful part of the N-Sink approach is not just the calculation of potential N removal for individual components of the landscape, it is the ability to summarize cumulative removal along the length of a specified flowpath. Starting from any specified location the flowpath on land is generated from a flow direction grid.  Once that flowpath intersects the surface water network, flow is determined by flow along the NHD Plus surface water network. With a flowpath generated the cumulative N removal along that flowpath can be calculated with the `nsink_summarize_flowpath()` function, taking the flowpath and removal as input. A data frame is returned with each segment identified by type of sink, if present (i.e., "Hydric", "Stream", "Lake/Pond", or "No Removal"), the percent removal associated with that segment, and cumulative removal.  Total cumulative removal is 100 - minimum of the `n_out` column. That is, `n_out` keeps track of the percent of N leaving each subsequent downstream segment (Table 2).

```{r summarize-it, message = FALSE, include = FALSE}
library(sf)
## Linking to GEOS 3.8.1, GDAL 3.1.4, PROJ 6.3.1
# Starting point

@@

```
```{r show-summary, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_a, caption = "Table 2: Summary of N removal along specified flowpath A (See Figure 4)")
```
## Static maps

@@

```{r static-maps, eval=TRUE, include = FALSE}
niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 450)
```

@@

```{r plot1, fig.cap = "Figure 1: Nitrogen Removal Efficiency (range 0 to 1) for the Niantic River HUC-12, showing the estimated percentage of N removed within three types of landscape N sinks - wetlands, streams, lakes/ponds/reservoirs."}

nsink_plot(niantic_static_maps, map = "removal")
```
```{r plot2, fig.cap = "Figure 2: Nitrogen Transport Index (range 0 to 100%) for the Niantic River HUC-12, showing the estimated percentage of N originating at a given location within a watershed that is expected to reach downstream receiving waters, such as a coastal embayment."}

nsink_plot(niantic_static_maps, map = "transport")
```
```{r plot3, fig.cap = "Figure 3: Nitrogen Delivery Index (range 0 to 100%) for the Niantic River HUC-12, showing a relative measure of the estimated N being transported from a location within a mapped watershed to receiving waters. N loading estimates are based on land cover classes, making use of peer-reviewed published data, and normalized, with a range of 0 to 1. These N loading factors are then multiplied by the N Transport Efficiency (range 0 to 100) from a given location within a watershed, to arrive at the NDI."}

@@

In estimating N removal occurring in the terrestrial portion of the flowpath, we originally focused on riparian wetlands, using SSURGO mapped hydric soils to identify this landscape sink. With the latest version of SSURGO we include all hydric soils in the catchment, except those identified as "subaqueous". Each soil mapping unit (SMU) that has a positive hydric status also has an estimate of %hydric for that SMU. We then estimate N removal within that SMU as 80% of the %hydric. Each raster has one SMU associated with it and raster cell size is 30m X 30m. A 30-m wide wetland is estimated by Mayer et al. [-@mayer2007meta] as having 80% N removal. In the current version N removal is cumulative as a flowpath intercepts hydric soils in any cell before encountering surface water.

## Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD data (e.g. time of travel) associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system. We examined seven HUC-12s within southern New England, with a range of land cover and soils, for the extent of off-network hydrology. None had off network stream or canal segments, while six of the seven had off-network waterbodies. These off-network waterbodies represented from 2 to 13% of total lake area. We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. A subset of these off-network stream segments are identified by the NHD dataset as ditches and canals. These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment due to the level of uncertainty regarding residence time in these groundwater-fed lakes and ponds [@hampton2019residence].

## Caveats
N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated. In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage bypasses the processes within wetland ecosystems, the removal of N in riparian zones is likely to be overstated. The variation in loading associated with different agricultural practices, density of unsewered development and inputs from wastewater treatment plants are not encompassed in the default loading index. Thus the numeric outputs of the tool focus on metrics suitable for planning purposes, such as percent removal of N from source to receiving water and relative intensity of loading from different locations. It is not intended to replace high-resolution models that rely on long term weather records and incorporate detailed spatial data on N loading, within-field transport or variable source hydrology (e.g., Easton [-@easton2008re]). Instead, N-Sink combines widely available datasets for waterway networks, soils and land cover with nitrogen dynamics developed for these datasets to highlight major sources and sinks of nitrogen within a watershed context.

@@

Users can use the following to interactively identify a starting point for a flowpath of interest. 
```{r case1, echo = TRUE, eval = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072
# Select location on map for starting point using cursor and click
pt <- unlist(locator(n = 1))
# Convert to sf POINT
start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
# Generate the flowpath
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
# Summarize N removal along the flowpath
niantic_removal <- nsink_calc_removal(niantic_data)
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) # table of removal within flowpath segments
```

Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters. This example illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Figure 4). The difference is due to the different N sinks encountered along the way to the estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3). These two currently undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet. If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.

```{r plot4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Figure 4: Flowpaths originating from two nearby locations within the Niantic River HUC-12 with differing N removal estimates. Hydrofeatures are pictured on the Nitrogen Transport Index map."}

@@

niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
                     "#F9C159", "#F99844", "#F56A2E","#EF2820"))
huc <- niantic_data$huc
streams <- st_intersection(huc, niantic_data$streams)
plot(streams, add = TRUE, col = "dodgerblue")
plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
ab_plot <- st_sf(st_sfc(ab), crs = aea)
plot(ab_plot, add = TRUE, pch = 16)
text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
```

```{r compare-flowpaths, include = TRUE, echo = FALSE}
knitr::kable(niantic_fp_removal_b, caption = "Table 3: Summary of N removal along specified flowpath B")
```

@@

The maps produced by N-Sink can be helpful in prioritizing source controls, sink protection and/or restoration. For example, many wetlands are considered "protected" by federal, state or local laws and regulations; but as occurs with many types of protected areas [@neal2018coastal] they are frequently altered through variances granted by local commissions or committees charged with zoning decisions. By better understanding the important role that N sinks play in the protection of estuaries, local agencies have a better chance of achieving protection of downstream receiving waters. Variances granted on a case-by-case basis can quickly undermine local water quality goals. Land conservancies and other NGOs interested in identifying parcels of high value for protecting downstream waters from N loading can also use these maps to prioritize land acquisition or wetland restoration projects.

@@

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two developed areas."}
box <- c(1949902, 1953930, 2288202, 2292262)
e <- extent(box)
legend_nlcd <- pal_nlcd()
case2 <- crop(niantic_data$nlcd, e)
vals <- unique(case2@data@values)
leg <- legend_nlcd[legend_nlcd$ID %in% vals,]
case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))

my_key <- list(rectangles=list(col = leg$Color),
              text=list(lab=leg$Class),
              space='left',
              columns=1,
              size=2,
              cex=.6)
colors <- legend_nlcd$Color
plot(case2, col = legend_nlcd$Color)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
```
```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
plot(case2, col = legend_nlcd$Color, alpha = 0.6, legend = FALSE)
plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
